---
title: "COVID 19 Analysis"
date: ""
output: pdf_document
---

#### Required Packages

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(usmap)
```

#### Part 1 - Basic Exploration of US Data

The New York Times (the Times) has aggregated reported COVID-19 data from state and local governments and health departments since 2020 and provides public access through a repository on GitHub. One of the data sets provided by the Times is county-level data for cumulative cases and deaths each day. This will be your primary data set for the first two parts of your analysis. 

County-level COVID data from 2020, 2021, and 2022 has been imported below. Each row of data reports the cumulative number of cases and deaths for a specific county each day. A FIPS code, a standard geographic identifier, is also provided which you will use in Part 2 to construct a map visualization at the county level for a state. 

Additionally, county-level population estimates reported by the US Census Bureau has been imported as well. You will use these estimates to caluclate statistics per 100,000 people. 

```{r import-nyt-data}
# Import New York Times COVID-19 data
# Import Population Estimates from US Census Bureau 

us_counties_2020 <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties-2020.csv")
us_counties_2021 <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties-2021.csv")
us_counties_2022 <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties-2022.csv")

us_population_estimates <- read_csv("fips_population_estimates.csv")
```

##### Question 1 

Your first task is to combine and tidy the 2020, 2021, and 2022 COVID data sets and find the total deaths and cases for each day since March 15, 2020 (2020-03-15). The data sets provided from the NY Times also includes statistics from Puerto Rico, a US territory. You may remove these observations from the data as they will not be needed for your analysis. Once you have tidied the data, find the total COVID-19 cases and deaths since March 15, 2020. Write a sentence or two after the code block communicating your results. Use inline code to include the `max_date`, `us_total_cases`, and `us_total_deaths` variables. To write inline code use `r `. 

```{r p1q1-response}

us_counties_2020
us_counties_2021
us_counties_2022
us_population_estimates

df_combined1<-bind_rows(us_counties_2020,us_counties_2021)
df_combined<-bind_rows(df_combined1,us_counties_2022)
df_combined


new_df <- df_combined[df_combined["state"]!="Puerto Rico",]
new_df

total_cases<-new_df%>%
  group_by(date)%>%
  summarize(
    total_deaths= sum(deaths),
    total_cases= sum(cases)
  )%>%
  filter(date>="2020-03-15")
total_cases
```

```{r p1q1-sample}

# Your output should look similar to the following tibble: 
#
#   A tibble: 657 x 3
#       date          total_deaths   total_cases
#      <date>            <dbl>         <dbl>
#   1 2020-03-15           68          3595
#   2 2020-03-16           91          4502
#   3 2020-03-17          117          5901
#   4 2020-03-18          162          8345
#   5 2020-03-19          212         12387
#   6 2020-03-20          277         17998
#   7 2020-03-21          359         24507
#   8 2020-03-22          457         33050
#   9 2020-03-23          577         43474
#  10 2020-03-24          783         53899
# ... with 647 more rows
#

```

##### Question 2 

Create a visualization for the total number of deaths and cases in the US since March 15, 2020. Before you create your visualization, review the types of plots you can create using the ggplot2 library and think about which plots would be effective in communicating your results. After you have created your visualization, write a few sentences describing your visualization. How could the plot be interpreted? Could it be misleading? 

```{r p1q2-response}

# Create a visualization for the total number of US cases and deaths since March 15, 2020. 
#
ggplot(total_cases,aes(x=date))+
  geom_line(aes(y=total_cases),color="blue")

ggplot(total_cases,aes(x=date))+
  geom_line(aes(y=total_deaths),color="red")

```

-- Communicate your methodology, results, and interpretation here -- 

##### Question 3

While it is important to know the total deaths and cases throughout the COVID-19 pandemic, it is also important for local and state health officials to know the the number of new cases and deaths each day to understand how rapidly the virus is spreading. Using the table you created in Question 1, calculate the number of new deaths and cases each day and a seven-day average of new deaths and cases. Once you have organized your data, find the days that saw the largest number of new cases and deaths. Write a sentence or two after the code block communicating your results.

```{r p1q3-response}

delta_deaths_1<-total_cases["total_deaths"]-lag(total_cases["total_deaths"],n=1)
delta_cases_1<-total_cases["total_cases"]-lag(total_cases["total_cases"],n=1)
total_cases["delta_deaths_1"]<-delta_deaths_1
total_cases["delta_cases_1"]<-delta_cases_1
total_cases$delta_deaths_1[is.na(total_cases$delta_deaths_1)]<-0
total_cases$delta_cases_1[is.na(total_cases$delta_cases_1)]<-0

total_cases<-total_cases %>%
  mutate(delta_deaths_7=((lag(delta_deaths_1,n=0) +
  lag(delta_deaths_1,n=2)+lag(delta_deaths_1,n=3)+lag(delta_deaths_1,n=4)+
  lag(delta_deaths_1,n=5)+lag(delta_deaths_1,n=6)+lag(delta_deaths_1,n=1))/7))

total_cases<-total_cases %>%
  mutate(delta_cases_7=((lag(delta_cases_1,n=0) +
  lag(delta_cases_1,n=2) +lag(delta_cases_1,n=3)+lag(delta_cases_1,n=4) +
  lag(delta_cases_1,n=5) +lag(delta_cases_1,n=6)+lag(delta_cases_1,n=1))/7))

total_cases[7,"delta_deaths_7"]<-NA
total_cases[7,"delta_cases_7"]<-NA

total_cases

max_new_deaths <- max(total_cases$delta_deaths_1)
max_new_deaths
max_new_deaths_date <- total_cases[total_cases$delta_deaths_1 == max_new_deaths,1]
max_new_deaths_date
max_new_cases <- max(total_cases$delta_cases_1)
max_new_cases
max_new_cases_date <- total_cases[total_cases$delta_cases_1 == max_new_cases,1]
max_new_cases_date
```

```{r p1q3-sample}

#  Your output should look similar to the following tibble: 
#  
#  date
#  total_deaths     >  the cumulative number of deaths up to and including the associated date
#  total_cases      >  the cumulative number of cases up to and including the associated date
#  delta_deaths_1   >  the number of new deaths since the previous day
#  delta_cases_1    >  the number of new cases since the previous day
#  delta_deaths_7   >  the average number of deaths in a seven-day period
#  delta_cases_7    >  the average number of cases in a seven-day period
#==
#  A tibble: 813 x 7
#    date           total_deaths   total_cases   delta_deaths_1    delta_cases_1  delta_deaths_7  delta_cases_7
#    <date>            <dbl>          <dbl>         <dbl>               <dbl>        <dbl>            <dbl>
#  1 2020-03-15           68          3600            0                     0         NA                NA 
#  2 2020-03-16           91          4507           23                   907         NA                NA 
#  3 2020-03-17          117          5906           26                  1399         NA                NA 
#  4 2020-03-18          162          8350           45                  2444         NA                NA 
#  5 2020-03-19          212         12393           50                  4043         NA                NA 
#  6 2020-03-20          277         18012           65                  5619         NA                NA 
#  7 2020-03-21          360         24528           83                  6516         NA                NA 
#  8 2020-03-22          458         33073           98                  8545       55.7              4210.
#  9 2020-03-23          579         43505          121                 10432       69.7              5571.
# 10 2020-03-24          785         53938          206                 10433       95.4              6862.
# ... with 803 more rows



```

-- Communicate your methodology, results, and interpretation here -- 

##### Question 4


```{r p1q4-response}

# Create a new table, based on the table from Question 3, and calculate the number of new deaths and cases per 100,000 people each day and a seven day average of new deaths and cases per 100,000 people. 

# Hint: To calculate per 100,000 people, first tidy the population estimates data and calculate the US population in 2020 and 2021. Then, you will need to divide  each statistic by the estimated population and then multiply by 100,000.
#
# Hint: look at the help documentation for grepl() and case_when() to divide the averages by the US population for each year. 
# For example, take the simple tibble, t_new: 
#
#     x     y   
#   <int> <chr>  
#     1     a   
#     2     b   
#     3     a    
#     4     b  
#     5     a  
#     6     b
#
#
# To add a column, z, that is dependent on the value in y, you could: 
#
# t_new %>%  
#   mutate(z = case_when(grepl("a", y) ~ "not b", 
#                        grepl("b", y) ~ "not a"))
#

total_pop_per_year<-us_population_estimates %>%
  group_by(Year)%>%
  summarize(total_pop= sum(Estimate))
total_pop_per_year

pop_in_2020<-total_pop_per_year$total_pop[1]
pop_in_2020
pop_in_2021<-total_pop_per_year$total_pop[2]
pop_in_2021

total_cases_100k<-total_cases %>%
  filter(date<="2022-01-01")

total_cases_100k %>%
  mutate(
    total_deaths = case_when(
      grepl("2020", date) ~ total_deaths/pop_in_2020*100000,
      grepl("2021", date) ~ total_deaths/pop_in_2021*100000
    )
  )

total_cases_100k<-total_cases_100k%>%
  mutate(
    total_cases= case_when(
      grepl("2020",date) ~total_cases/pop_in_2020*100000,
      grepl("2021",date) ~total_cases/pop_in_2021*100000
  )
)

total_cases_100k<-total_cases_100k%>%
  mutate(
    delta_deaths_1 = case_when(
      grepl("2020",date) ~delta_deaths_1/pop_in_2020*100000,
      grepl("2021",date) ~delta_deaths_1/pop_in_2021*100000
  )
)
total_cases_100k<-total_cases_100k%>%
  mutate(
    delta_cases_1= case_when(
      grepl("2020",date) ~delta_cases_1/pop_in_2020*100000,
      grepl("2021",date) ~delta_cases_1/pop_in_2021*100000
    )
  )

total_cases_100k<-total_cases_100k%>%
  mutate(
    delta_deaths_7= case_when(
      grepl("2020",date) ~delta_deaths_7/pop_in_2020*100000,
      grepl("2021",date) ~delta_deaths_7/pop_in_2021*100000
    )
)

total_cases_100k<-total_cases_100k%>%
  mutate(
    delta_cases_7= case_when(
      grepl("2020",date) ~delta_cases_7/pop_in_2020*100000,
      grepl("2021",date) ~delta_cases_7/pop_in_2021*100000
  )
)
total_cases_100k

```

```{r p1q4-sample}

#  Your output should look similar to the following tibble: 
#  
#  date
#  total_deaths     >  the cumulative number of deaths up to and including the associated date
#  total_cases      >  the cumulative number of cases up to and including the associated date
#  delta_deaths_1   >  the number of new deaths since the previous day
#  delta_cases_1    >  the number of new cases since the previous day
#  delta_deaths_7   >  the average number of deaths in a seven-day period
#  delta_cases_7    >  the average number of cases in a seven-day period
#==
#  A tibble: 657 x 7
#       date         total_deaths   total_cases   delta_deaths_1   delta_cases_1 delta_deaths_7 delta_cases_7
#      <date>            <dbl>         <dbl>          <dbl>           <dbl>           <dbl>         <dbl>
#   1 2020-03-15         0.0205        1.08                0              0              NA           NA   
#   2 2020-03-16         0.0275        1.36          0.00694          0.274              NA           NA   
#   3 2020-03-17         0.0353        1.78          0.00784          0.422              NA           NA   
#   4 2020-03-18         0.0489        2.52           0.0136          0.737              NA           NA   
#   5 2020-03-19         0.0640        3.74           0.0151           1.22              NA           NA   
#   6 2020-03-20         0.0836        5.43           0.0196           1.69              NA           NA   
#   7 2020-03-21         0.108         7.39           0.0247           1.96              NA           NA   
#   8 2020-03-22         0.138         9.97           0.0296           2.58          0.0168         1.27
#   9 2020-03-23         0.174         13.1           0.0362           3.14          0.0209         1.68
#  10 2020-03-24         0.236         16.3           0.0621           3.14          0.0287         2.07
```

-- Communicate your methodology, results, and interpretation here -- 

#### Question 5

```{r p1q5-response}

# Create a visualization to compare the seven-day average cases and deaths per 100,000 people. 

ggplot(total_cases_100k, aes(x=date, y=delta_deaths_7)) +
  geom_line(color = "red")

ggplot(total_cases_100k, aes(x=date, y=delta_cases_7)) +
  geom_line(color = "blue")
  

```

-- Communicate your methodology, results, and interpretation here -- 

##### Part 2 - US State Comparison

While understanding the trends on a national level can be helpful in understanding how COVID-19 impacted the United States, it is important to remember that the virus arrived in the United States at different times. For the next part of your analysis, you will begin to look at COVID related deaths and cases at the state and county-levels.

##### Question 1

Your first task in Part 2 is to determine the top 10 states in terms of total deaths and cases between March 15, 2020, and December 31, 2021.

Once you have both lists, briefly describe your methodology and your results.

```{r p2q1-response}

# Determine the top 10 states in terms of total deaths and cases between March 15, 2020, and December 31, 2021. To do this, transform your combined COVID-19 data to summarize total deaths and cases by state up to December 31, 2021. 

updated_df <- new_df %>%
  filter(date == "2021-12-31") %>%
  group_by(state) %>%
  summarize(
    total_deaths = sum(deaths, na.rm = TRUE),
    total_cases = sum(cases, na.rm = TRUE)
  ) %>%
  arrange(desc(total_deaths), desc(total_cases))

updated_df$date = as.Date("2021-12-31")

updated_df

updated_df <- updated_df %>%
  select(state, date, total_deaths, total_cases)

updated_df

print("California has had the most number of cases between March 15, 2020, and December 31, 2021.")
```

```{r p2q1-sample}

# Your transformed data should look similar to the following tibble:  
#
# A tibble: 51 x 4
#     state                  date       total_deaths  total_cases
#     <chr>                 <date>          <dbl>       <dbl>
#  1 California           2021-12-31        76709      5515613
#  2 Texas                2021-12-31        76062      4574881
#  3 Florida              2021-12-31        62504      4166392
#  4 New York             2021-12-31        58993      3473970
#  5 Illinois             2021-12-31        31017      2154058
#  6 Pennsylvania         2021-12-31        36705      2036424
#  7 Ohio                 2021-12-31        29447      2016095
#  8 Georgia              2021-12-31        30283      1798497
#  9 Michigan             2021-12-31        28984      1706355
# 10 North Carolina       2021-12-31        19436      1685504
# ... with 41 more rows
```

-- Communicate your methodology, results, and interpretation here -- 

##### Question 2

Determine the top 10 states in terms of deaths per 100,000 people and cases per 100,000 people between March 15, 2020, and December 31, 2021.

Once you have both lists, briefly describe your methodology and your results. Do you expect the lists to be different than the one produced in Question 1? Which method, total or per 100,000 people, is a better method for reporting the statistics? 

```{r p2q2-response}

# Determine the top 10 states in terms of deaths and cases per 100,000 people between March 15, 2020, and December 31, 2021. You should first tidy and transform the population estimates to include population totals by state. Use your relational data verbs (e.g. full_join()) to join the population estimates with the cases and death statistics using the state name as a key. Then, use case_when() and grepl() to add a population column to your table that only includes the estimated population for the associated year. Finally, mutate your table to calculate deaths and cases per 100,000 people and summarize by state. 

updated_df 

state_pop <- us_population_estimates %>%
  group_by(STNAME, Year) %>%
  summarize(Estimate = sum(Estimate))

updated_df_per_100k <- updated_df %>%
  left_join(state_pop, by = c("state" = "STNAME")) %>%
  mutate(deaths_per_100k = total_deaths/Estimate*100000,
         cases_per_100k = total_cases/Estimate*100000) %>%
  select(state, date, Year, deaths_per_100k, cases_per_100k) %>%
  filter(Year == 2021) %>%
  select(state, date, deaths_per_100k, cases_per_100k) %>%
  arrange(desc(cases_per_100k))

updated_df_per_100k

print("A key finding is that California may have the most cases, but
      in terms of per 100k population it is not one of the most. This
      tells us that a key factor why it had such high total cases and 
      deaths is mainly due to such a high population in the year 2021.")
```

```{r p2q2-sample}

# Your transformed data should look similar to the following tibble:
#
# A tibble: 51 x 4
#     state                   date           deaths_per_100k  cases_per_100k
#     <chr>                  <date>               <dbl>          <dbl>
#  1 North Dakota          2021-12-31             265.           22482.
#  2 Alaska                2021-12-31             130.           21310.
#  3 Rhode Island          2021-12-31             280.           21093.
#  4 South Dakota          2021-12-31             278.           20014.
#  5 Wyoming               2021-12-31             264.           19979.
#  6 Tennessee             2021-12-31             296.           19783.
#  7 Kentucky              2021-12-31             269.           19173.
#  8 Florida               2021-12-31             287.           19128.
#  9 Utah                  2021-12-31             113.           19088.
# 10 Wisconsin             2021-12-31             190.           19008.
# ... with 41 more rows
```

-- Communicate your methodology, results, and interpretation here -- 

##### Question 3

Now, select a state and calculate the seven-day averages for new cases and deaths per 100,000 people. Once you have calculated the averages, create a visualization using ggplot2 to represent the data. 

```{r p2q3-response}

# Select a state and then filter by state and date range your data from Question 1. Calculate the seven-day average following the same procedure as Part 1. 

the_state = "Colorado"

df_for_2020 <- new_df %>%
  filter(state == the_state & date >= "2020-03-15" & date < "2021-01-01") %>%
  group_by(date) %>%
  summarize(
    total_deaths = sum(deaths),
    total_cases = sum(cases)
  ) %>%
  mutate(state = the_state) %>%
  select(state, date, total_deaths, total_cases)

state_pop_for_2020 <- state_pop %>%
  filter(Year == 2020) %>%
  filter(STNAME == the_state) %>%
  select(Estimate) %>%
  rename(population = Estimate)

df_for_2020 
state_pop_for_2020

final_df_2020 <- df_for_2020 %>%
  left_join(state_pop_for_2020, by = c("state" = "STNAME"))

final_df_2020

#-------------------------------------------------------------------

df_for_2021 <- new_df %>%
  filter(state == the_state & date >= "2021-01-01" & date < "2022-01-01") %>%
  group_by(date) %>%
  summarize(
    total_deaths = sum(deaths),
    total_cases = sum(cases)
  ) %>%
  mutate(state = the_state) %>%
  select(state, date, total_deaths, total_cases)

state_pop_for_2021 <- state_pop %>%
  filter(Year == 2021) %>%
  filter(STNAME == the_state) %>%
  select(Estimate) %>%
  rename(population = Estimate)

df_for_2021 
state_pop_for_2021

final_df_2021 <- df_for_2021 %>%
  left_join(state_pop_for_2021, by = c("state" = "STNAME"))

final_df_2021

#---------------------------------------------------------------------

final_df <- rbind(final_df_2020, final_df_2021)

state_7_avg <- final_df %>%
  mutate(
    deaths_per_100k = total_deaths/population*100000,
    cases_per_100k = total_cases/population*100000,
    new_death = deaths_per_100k - lag(deaths_per_100k, n=1),
    new_case = cases_per_100k - lag(cases_per_100k, n=1)
  ) %>%
  mutate(
    deaths_7_day = ((new_death) + lag(new_death, n=1) + lag(new_death, n=2) +
      lag(new_death, n=3) + lag(new_death, n=4) + lag(new_death, n=5) +
      lag(new_death, n=6))/7,
    cases_7_day = ((new_case) + lag(new_case, n=1) + lag(new_case, n=2) +
      lag(new_case, n=3) + lag(new_case, n=4) + lag(new_case, n=5) +
      lag(new_case, n=6))/7 
  ) %>%
  select(-new_death, -new_case)

state_7_avg

print("We see initially cases and deaths were low but the daily change
      increases.")
```

```{r p2q3-sample}

# Your transformed data should look similar to the following tibble: 
#
# A tibble: 656 × 9
#     state    date       total_deaths total_cases population deaths_per_100k cases_per_100k deaths_7_day  cases_7_day
#     <chr>    <date>            <dbl>    <dbl>      <dbl>         <dbl>          <dbl>        <dbl>         <dbl>
#  1 Colorado 2020-03-15            2      136      5784308        0.0346          2.35         NA            NA   
#  2 Colorado 2020-03-16            2      161      5784308        0.0346          2.78         NA            NA   
#  3 Colorado 2020-03-17            3      183      5784308        0.0519          3.16         NA            NA   
#  4 Colorado 2020-03-18            3      216      5784308        0.0519          3.73         NA            NA   
#  5 Colorado 2020-03-19            5      278      5784308        0.0864          4.81         NA            NA   
#  6 Colorado 2020-03-20            5      364      5784308        0.0864          6.29         NA            NA   
#  7 Colorado 2020-03-21            6      475      5784308        0.104           8.21         NA            NA   
#  8 Colorado 2020-03-22            7      591      5784308        0.121           10.2       0.0123         1.12
#  9 Colorado 2020-03-23           10      721      5784308        0.173           12.5       0.0198         1.38
# 10 Colorado 2020-03-24           11      912      5784308        0.190           15.8       0.0198         1.80
# … with 646 more rows
```

-- Communicate your methodology, results, and interpretation here -- 

##### Question 4

Using the same state, identify the top 5 counties in terms of deaths and cases per 100,000 people. 

```{r p2q4-response}

# Using the same state as Question 2, filter your state and date range from the combined data set from Part 1 and summarize cases and deaths. Produce two lists arranged by deaths and cases. When transforming the data, be sure to include the "fips" column as you will need this to complete Question 5. 

the_state = "Colorado"

colorado <- new_df %>%
  filter(state == the_state & date == "2021-12-20") %>%
  mutate(total_deaths = deaths,
         total_cases = cases) %>%
  arrange(desc(total_cases)) %>%
  select(county, date, fips, total_deaths, total_cases)

colorado

print("In the state of Colorado, El Paso has the most number of cases and deaths from COVID.")
```

```{r p2q4-sample}

# Your transformed data should be similar to the following tibbles: 
#
# Arranged by deaths: 
# A tibble: 64 × 4
#     county       date       fips    total_deaths   total_cases
#     <chr>       <date>      <chr>       <dbl>         <dbl>
#  1 El Paso    2021-12-20    08041       1355         119772
#  2 Denver     2021-12-20    08031       1065         106747
#  3 Jefferson  2021-12-20    08059       1061         76732
#  4 Adams      2021-12-20    08001       1057         90476
#  5 Arapahoe   2021-12-20    08005       1046         95769
#  6 Pueblo     2021-12-20    08101        643         30739
#  7 Weld       2021-12-20    08123        569         55599
#  8 Mesa       2021-12-20    08077        445         29542
#  9 Larimer    2021-12-20    08069        393         47444
# 10 Douglas    2021-12-20    08035        361         48740
# … with 54 more rows
#
#
# Arranged by cases: 
# A tibble: 64 × 4
#     county       date       fips   total_deaths   total_cases
#     <chr>       <date>      <chr>     <dbl>         <dbl>
#  1 El Paso    2021-12-20    08041     1355         119772
#  2 Denver     2021-12-20    08031     1065         106747
#  3 Arapahoe   2021-12-20    08005     1046          95769
#  4 Adams      2021-12-20    08001     1057          90476
#  5 Jefferson  2021-12-20    08059     1061          76732
#  6 Weld       2021-12-20    08123      569          55599
#  7 Douglas    2021-12-20    08035      361          48740
#  8 Larimer    2021-12-20    08069      393          47444
#  9 Boulder    2021-12-20    08013      323          36754
# 10 Pueblo     2021-12-20    08101      643          30739
# … with 54 more rows
```

-- Communicate your methodology, results, and interpretation here -- 

##### Question 5

Modify the code below for the map projection to plot county-level deaths and cases per 100,000 people for your state. 

```{r p2q5-response}
state_pop_for_2021$population

colorado_100k <- colorado %>% 
  mutate(
    total_deaths_100k = total_deaths/state_pop_for_2021$population*100000,
    total_cases_100k = total_cases/state_pop_for_2021$population*100000
  )
colorado_100k
```

```{r p2q5-sample}
# Copy and modify the code below for your state.
#
# plot_usmap arguments: 
#   regions: can be one of ("states", "state", "counties", "county"). The default is "states"
#   include: The regions to include in the resulting map. If regions is "states"/"state", the value can be either a state name, abbreviation or FIPS code. For counties, the FIPS must be provided as there can be multiple counties with the same name. 
#   data: values to plot on the map
#   values: the name of the column that contains the values to be associated with a given region. 
#   color: the map outline color.
#
# Reference the plot_usmap documentation for further information using ?plot_usmap

plot_usmap(regions = "county", include="CO", data = colorado_100k, values = "total_deaths_100k", color = "blue") +
  scale_fill_continuous(low = "white", high = "blue", name = "Deaths per 100,000")

```

-- Communicate your methodology, results, and interpretation here -- 

##### Question 6

Finally, select three other states and calculate the seven-day averages for new deaths and cases per 100,000 people for between March 15, 2020, and December 31, 2021. 


```{r p2q6-response}

the_state = "Colorado"

df_for_2020 <- new_df %>%
  filter(state == the_state & date >= "2020-03-15" & date < "2021-01-01") %>%
  group_by(date) %>%
  summarize(
    total_deaths = sum(deaths),
    total_cases = sum(cases)
  ) %>%
  mutate(state = the_state) %>%
  select(state, date, total_deaths, total_cases)

state_pop_for_2020 <- state_pop %>%
  filter(Year == 2020) %>%
  filter(STNAME == the_state) %>%
  select(Estimate) %>%
  rename(population = Estimate)

df_for_2020 
state_pop_for_2020

final_df_2020 <- df_for_2020 %>%
  left_join(state_pop_for_2020, by = c("state" = "STNAME"))

final_df_2020

#-------------------------------------------------------------------

df_for_2021 <- new_df %>%
  filter(state == the_state & date >= "2021-01-01" & date < "2022-01-01") %>%
  group_by(date) %>%
  summarize(
    total_deaths = sum(deaths),
    total_cases = sum(cases)
  ) %>%
  mutate(state = the_state) %>%
  select(state, date, total_deaths, total_cases)

state_pop_for_2021 <- state_pop %>%
  filter(Year == 2021) %>%
  filter(STNAME == the_state) %>%
  select(Estimate) %>%
  rename(population = Estimate)

df_for_2021 
state_pop_for_2021

final_df_2021 <- df_for_2021 %>%
  left_join(state_pop_for_2021, by = c("state" = "STNAME"))

final_df_2021

#---------------------------------------------------------------------

final_df <- rbind(final_df_2020, final_df_2021)

state_7_avg <- final_df %>%
  mutate(
    deaths_per_100k = total_deaths/population*100000,
    cases_per_100k = total_cases/population*100000,
    new_death = deaths_per_100k - lag(deaths_per_100k, n=1),
    new_case = cases_per_100k - lag(cases_per_100k, n=1)
  ) %>%
  mutate(
    deaths_7_day = ((new_death) + lag(new_death, n=1) + lag(new_death, n=2) +
      lag(new_death, n=3) + lag(new_death, n=4) + lag(new_death, n=5) +
      lag(new_death, n=6))/7,
    cases_7_day = ((new_case) + lag(new_case, n=1) + lag(new_case, n=2) +
      lag(new_case, n=3) + lag(new_case, n=4) + lag(new_case, n=5) +
      lag(new_case, n=6))/7 
  ) %>%
  select(-new_death, -new_case)

Colorado_7 <- state_7_avg %>%
  select(state, date, deaths_7_day, cases_7_day)

Colorado_7

#---------------------------------------------------------------------------
#---------------------------------------------------------------------------
#---------------------------------------------------------------------------
#---------------------------------------------------------------------------
#---------------------------------------------------------------------------

the_state = "California"

df_for_2020 <- new_df %>%
  filter(state == the_state & date >= "2020-03-15" & date < "2021-01-01") %>%
  group_by(date) %>%
  summarize(
    total_deaths = sum(deaths),
    total_cases = sum(cases)
  ) %>%
  mutate(state = the_state) %>%
  select(state, date, total_deaths, total_cases)

state_pop_for_2020 <- state_pop %>%
  filter(Year == 2020) %>%
  filter(STNAME == the_state) %>%
  select(Estimate) %>%
  rename(population = Estimate)

df_for_2020 
state_pop_for_2020

final_df_2020 <- df_for_2020 %>%
  left_join(state_pop_for_2020, by = c("state" = "STNAME"))

final_df_2020

#-------------------------------------------------------------------

df_for_2021 <- new_df %>%
  filter(state == the_state & date >= "2021-01-01" & date < "2022-01-01") %>%
  group_by(date) %>%
  summarize(
    total_deaths = sum(deaths),
    total_cases = sum(cases)
  ) %>%
  mutate(state = the_state) %>%
  select(state, date, total_deaths, total_cases)

state_pop_for_2021 <- state_pop %>%
  filter(Year == 2021) %>%
  filter(STNAME == the_state) %>%
  select(Estimate) %>%
  rename(population = Estimate)

df_for_2021 
state_pop_for_2021

final_df_2021 <- df_for_2021 %>%
  left_join(state_pop_for_2021, by = c("state" = "STNAME"))

final_df_2021

#---------------------------------------------------------------------

final_df <- rbind(final_df_2020, final_df_2021)

state_7_avg <- final_df %>%
  mutate(
    deaths_per_100k = total_deaths/population*100000,
    cases_per_100k = total_cases/population*100000,
    new_death = deaths_per_100k - lag(deaths_per_100k, n=1),
    new_case = cases_per_100k - lag(cases_per_100k, n=1)
  ) %>%
  mutate(
    deaths_7_day = ((new_death) + lag(new_death, n=1) + lag(new_death, n=2) +
      lag(new_death, n=3) + lag(new_death, n=4) + lag(new_death, n=5) +
      lag(new_death, n=6))/7,
    cases_7_day = ((new_case) + lag(new_case, n=1) + lag(new_case, n=2) +
      lag(new_case, n=3) + lag(new_case, n=4) + lag(new_case, n=5) +
      lag(new_case, n=6))/7 
  ) %>%
  select(-new_death, -new_case)

California_7 <- state_7_avg %>%
  select(state, date, deaths_7_day, cases_7_day)

California_7

#---------------------------------------------------------------------------
#---------------------------------------------------------------------------
#---------------------------------------------------------------------------
#---------------------------------------------------------------------------
#---------------------------------------------------------------------------

the_state = "Texas"

df_for_2020 <- new_df %>%
  filter(state == the_state & date >= "2020-03-15" & date < "2021-01-01") %>%
  group_by(date) %>%
  summarize(
    total_deaths = sum(deaths),
    total_cases = sum(cases)
  ) %>%
  mutate(state = the_state) %>%
  select(state, date, total_deaths, total_cases)

state_pop_for_2020 <- state_pop %>%
  filter(Year == 2020) %>%
  filter(STNAME == the_state) %>%
  select(Estimate) %>%
  rename(population = Estimate)

df_for_2020 
state_pop_for_2020

final_df_2020 <- df_for_2020 %>%
  left_join(state_pop_for_2020, by = c("state" = "STNAME"))

final_df_2020

#-------------------------------------------------------------------

df_for_2021 <- new_df %>%
  filter(state == the_state & date >= "2021-01-01" & date < "2022-01-01") %>%
  group_by(date) %>%
  summarize(
    total_deaths = sum(deaths),
    total_cases = sum(cases)
  ) %>%
  mutate(state = the_state) %>%
  select(state, date, total_deaths, total_cases)

state_pop_for_2021 <- state_pop %>%
  filter(Year == 2021) %>%
  filter(STNAME == the_state) %>%
  select(Estimate) %>%
  rename(population = Estimate)

df_for_2021 
state_pop_for_2021

final_df_2021 <- df_for_2021 %>%
  left_join(state_pop_for_2021, by = c("state" = "STNAME"))

final_df_2021

#---------------------------------------------------------------------

final_df <- rbind(final_df_2020, final_df_2021)

state_7_avg <- final_df %>%
  mutate(
    deaths_per_100k = total_deaths/population*100000,
    cases_per_100k = total_cases/population*100000,
    new_death = deaths_per_100k - lag(deaths_per_100k, n=1),
    new_case = cases_per_100k - lag(cases_per_100k, n=1)
  ) %>%
  mutate(
    deaths_7_day = ((new_death) + lag(new_death, n=1) + lag(new_death, n=2) +
      lag(new_death, n=3) + lag(new_death, n=4) + lag(new_death, n=5) +
      lag(new_death, n=6))/7,
    cases_7_day = ((new_case) + lag(new_case, n=1) + lag(new_case, n=2) +
      lag(new_case, n=3) + lag(new_case, n=4) + lag(new_case, n=5) +
      lag(new_case, n=6))/7 
  ) %>%
  select(-new_death, -new_case)

Texas_7 <- state_7_avg %>%
  select(state, date, deaths_7_day, cases_7_day)

Texas_7

#---------------------------------------------------------------------------
#---------------------------------------------------------------------------
#---------------------------------------------------------------------------
#---------------------------------------------------------------------------
#---------------------------------------------------------------------------

the_state = "Florida"

df_for_2020 <- new_df %>%
  filter(state == the_state & date >= "2020-03-15" & date < "2021-01-01") %>%
  group_by(date) %>%
  summarize(
    total_deaths = sum(deaths),
    total_cases = sum(cases)
  ) %>%
  mutate(state = the_state) %>%
  select(state, date, total_deaths, total_cases)

state_pop_for_2020 <- state_pop %>%
  filter(Year == 2020) %>%
  filter(STNAME == the_state) %>%
  select(Estimate) %>%
  rename(population = Estimate)

df_for_2020 
state_pop_for_2020

final_df_2020 <- df_for_2020 %>%
  left_join(state_pop_for_2020, by = c("state" = "STNAME"))

final_df_2020

#-------------------------------------------------------------------

df_for_2021 <- new_df %>%
  filter(state == the_state & date >= "2021-01-01" & date < "2022-01-01") %>%
  group_by(date) %>%
  summarize(
    total_deaths = sum(deaths),
    total_cases = sum(cases)
  ) %>%
  mutate(state = the_state) %>%
  select(state, date, total_deaths, total_cases)

state_pop_for_2021 <- state_pop %>%
  filter(Year == 2021) %>%
  filter(STNAME == the_state) %>%
  select(Estimate) %>%
  rename(population = Estimate)

df_for_2021 
state_pop_for_2021

final_df_2021 <- df_for_2021 %>%
  left_join(state_pop_for_2021, by = c("state" = "STNAME"))

final_df_2021

#---------------------------------------------------------------------

final_df <- rbind(final_df_2020, final_df_2021)

state_7_avg <- final_df %>%
  mutate(
    deaths_per_100k = total_deaths/population*100000,
    cases_per_100k = total_cases/population*100000,
    new_death = deaths_per_100k - lag(deaths_per_100k, n=1),
    new_case = cases_per_100k - lag(cases_per_100k, n=1)
  ) %>%
  mutate(
    deaths_7_day = ((new_death) + lag(new_death, n=1) + lag(new_death, n=2) +
      lag(new_death, n=3) + lag(new_death, n=4) + lag(new_death, n=5) +
      lag(new_death, n=6))/7,
    cases_7_day = ((new_case) + lag(new_case, n=1) + lag(new_case, n=2) +
      lag(new_case, n=3) + lag(new_case, n=4) + lag(new_case, n=5) +
      lag(new_case, n=6))/7 
  ) %>%
  select(-new_death, -new_case)

Florida_7 <- state_7_avg %>%
  select(state, date, deaths_7_day, cases_7_day)

Florida_7

#---------------------------------------------------------------------------
#---------------------------------------------------------------------------
#---------------------------------------------------------------------------
#---------------------------------------------------------------------------
#---------------------------------------------------------------------------

Colorado_7
California_7
Texas_7
Florida_7
```

-- Communicate your methodology, results, and interpretation here -- 

##### Question 7

Create a visualization comparing the seven-day averages for new deaths and cases per 100,000 people for the four states you selected. 

```{r p2q7-response}
fully_combined <- bind_rows(Colorado_7, California_7, Texas_7, Florida_7)
fully_combined

ggplot(fully_combined, aes(x = date, y = deaths_7_day)) +
  geom_line(aes(color = state)) +
  labs(
    title = "Deaths 7-day Average per 100k population of Various States"
  )

print("Key Findings:
      For California, the death rate was managed well, except for 2021 Q1.
      For Colorado, it had spikes in high death-rate in the start, Q4 of 2020, and Q4 of 2021.
      For Florida, it did a terrible job in managing death-rates in Q3 of 2021.
      Lastly, for Texas, it had 3 peaks of the same height: Q3 of 2020, Q1 of 2021, and Q3 of 2021 ")

ggplot(fully_combined, aes(x = date, y = cases_7_day)) +
  geom_line(aes(color = state)) +
  labs(
    title = "Cases 7-day Average per 100k population of Various States"
  )
```

-- Communicate your methodology, results, and interpretation here -- 

#### Part 3 - Global Comparison

```{r import-csse}
# Import global COVID-19 statistics aggregated by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University.
# Import global population estimates from the World Bank.

csse_global_deaths <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")
csse_global_cases <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")
csse_us_deaths <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")
csse_us_cases <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")

global_population_estimates <- read_csv("global_population_estimates.csv")
```

##### Question 1

Using the state you selected in Part 2 Question 2 compare the daily number of cases and deaths reported from the CSSE and NY Times. 

```{r p3q1-response}

# To compare your state data between the two data sets, you will first need to tidy the US CSSE death and cases data.
# Hint: Review the documentation for pivot_longer().

# ----------------------------NY TIMES----------------------------

new_df %>%
  filter(state == "Colorado" & date >= "2020-03-15" & date < "2022-01-01") %>%
  arrange(county) %>%
  select(fips, county, state, cases, deaths)

# ----------------------------CSSE US----------------------------

csse_us_cases <- csse_us_cases %>%
  select(FIPS, Admin2, Province_State, `1/22/20`:`10/6/22`)

csse_us_cases_longer <- csse_us_cases %>%
  pivot_longer(cols = contains("/"),
               names_to = "date",
               values_to = "cases") %>%
  mutate(
    date = as.Date(date, format = "%m/%d/%y")
  ) %>%
  filter(Province_State == "Colorado" & date >= "2020-03-15" & date < "2022-01-01")

csse_us_cases_longer

csse_us_deaths <- csse_us_deaths %>%
  select(FIPS, Admin2, Province_State, `1/22/20`:`10/6/22`)

csse_us_deaths_longer <- csse_us_deaths %>%
  pivot_longer(cols = contains("/"), names_to = "date", values_to = "deaths") %>%
  mutate(date = as.Date(date, format = "%m/%d/%y")) %>%
  filter(Province_State == "Colorado" & date >= "2020-03-15" & date < "2022-01-01")

csse_us_deaths_longer <- csse_us_deaths_longer %>%
  select(FIPS, deaths)

csse_us_colorado <- csse_us_cases_longer %>%
  left_join(csse_us_deaths_longer, by = c("FIPS" = "FIPS"))

csse_us_colorado <- csse_us_colorado %>%
  distinct(FIPS, Admin2, Province_State, date, .keep_all = TRUE)

csse_us_colorado

print("They are the same.")

# Once you have tidied your data, join the two CSSE US data sets to include cases and deaths in one table. 
# Finally, create two visualizations with one plotting the CSSE and NY Times cases and the other plotting the CSEE and NY Times deaths.

```

```{r p3q1-sample}

# Your tidied CSSE data for your selected state should look similar to the following tibble:
#
# A tibble: 43,362 × 6
#     fips  county  state        date      cases  deaths
#     <dbl> <chr>   <chr>       <date>     <dbl>  <dbl>
#  1  8001  Adams  Colorado   2020-03-15     6      0
#  2  8001  Adams  Colorado   2020-03-16     8      0
#  3  8001  Adams  Colorado   2020-03-17    10      0
#  4  8001  Adams  Colorado   2020-03-18    10      0
#  5  8001  Adams  Colorado   2020-03-19    10      0
#  6  8001  Adams  Colorado   2020-03-20    12      0
#  7  8001  Adams  Colorado   2020-03-21    14      0
#  8  8001  Adams  Colorado   2020-03-22    18      0
#  9  8001  Adams  Colorado   2020-03-23    25      0
# 10  8001  Adams  Colorado   2020-03-24    27      0
# … with 43,352 more rows

```

-- Communicate your methodology, results, and interpretation here -- 

##### Question 2 

Now that you have verified the data reported from the CSSE and NY Times are similar, combine the global and US CSSE data sets and identify the top 10 countries in terms of deaths and cases per 100,000 people between March 15, 2020, and December 31, 2021.

```{r p3q2-response}

# First, combine and tidy the CSSE death and cases data sets. You may wish to keep the two sets separate.
# Then, tidy the global population estimates. While tidying your data, remember to include columns that you will be able to use when joining the COVID-19 data. 
# You will notice that the population estimates data does not include every country reported in the CSSE data. When calculating statistics per 100,000 people, you will need to filter the CSSE data to only include countries that you have population estimates for. 


csse_global_cases_cleaned <- csse_global_cases %>%
  select(-"Province/State", -"Lat", -"Long") %>%
  mutate(Country_Name = `Country/Region`) %>%
  select(-"Country/Region") %>%
  select(Country_Name, everything())

csse_global_cases_longer <- csse_global_cases_cleaned %>%
  pivot_longer(cols = contains("/"),
               names_to = "date",
               values_to = "cases") %>%
  mutate(date = as.Date(date, format = "%m/%d/%y"))

csse_global_cases_longer

csse_global_country_cases <- csse_global_cases_longer %>%
  filter(date == "2021-12-31") 

csse_global_country_cases

#------------------------------------------

csse_global_deaths_cleaned <- csse_global_deaths %>%
  select(-"Province/State", -"Lat", -"Long") %>%
  mutate(Country_Name = `Country/Region`) %>%
  select(-"Country/Region") %>%
  select(Country_Name, everything())

csse_global_deaths_longer <- csse_global_deaths_cleaned %>%
  pivot_longer(cols = contains("/"),
               names_to = "date",
               values_to = "deaths") %>%
  mutate(date = as.Date(date, format = "%m/%d/%y"))

csse_global_deaths_longer %>%
  filter(date == "2021-12-31")

csse_global_country_deaths <- csse_global_deaths_longer %>%
  filter(date == "2021-12-31") %>%
  select(-"date")

csse_global_country_deaths 

#------------------------------------------

csse_global_combination <- csse_global_country_cases %>%
  left_join(csse_global_country_deaths, by = c("Country_Name" = "Country_Name")) 

csse_global_fixed <- csse_global_combination %>%
  group_by(Country_Name) %>%
  summarize(
    cases = sum(cases),
    deaths = sum(deaths)
  )

csse_global_fixed <- csse_global_fixed %>%
  mutate(
    date = "2021-12-31"
  ) %>%
  select(Country_Name, date, cases, deaths)

csse_global_fixed %>%
  arrange(desc(cases))

#------------------------------------------

global_population_cleaned <- global_population_estimates %>%
  rename(Country_Name = `Country Name`,
         `2021_Population` = `2021 [YR2021]`)

global_population_cleaned

global_population_fixed <- global_population_cleaned %>%
  select(Country_Name, `2021_Population`)
  
global_population_fixed
#------------------------------------------
csse_global_fixed
global_population_fixed

combined <- csse_global_fixed %>%
  left_join(global_population_fixed, by = c("Country_Name" = "Country_Name"))

combined

combined_cleaned <- combined %>%
  filter(!is.na(`2021_Population`))

combined_cleaned

combined_cleaned <- combined_cleaned %>%
  mutate(`2021_Population` = as.numeric(`2021_Population`))

the_100k <- combined_cleaned %>%
  mutate(
    cases_100k = round(cases/`2021_Population`*100000, 2),
    deaths_100k = round(deaths/`2021_Population`*100000, 2)
  )

the_100k

the_100k %>%
  arrange(desc(cases_100k))

the_100k %>%
  arrange(desc(deaths_100k))
```

-- Communicate your methodology, results, and interpretation here -- 

##### Question 3

Construct a visualization plotting the 10 countries in terms of deaths and cases per 100,000 people between March 15, 2020, and December 31, 2021. In designing your visualization keep the number of data you will be plotting in mind. You may wish to create two separate visualizations, one for deaths and another for cases. 

```{r p3q3-response}
the_cases <- the_100k %>%
  select(Country_Name, cases_100k) %>%
  arrange(desc(cases_100k)) %>%
  head(10)

the_deaths <- the_100k %>%
  select(Country_Name, deaths_100k) %>%
  arrange(desc(deaths_100k)) %>%
  head(10)

ggplot(the_cases, aes(x=Country_Name, y=cases_100k)) +
  geom_col()

ggplot(the_deaths, aes(x=Country_Name, y=deaths_100k)) +
  geom_col()
```

-- Communicate your methodology, results, and interpretation here -- 

##### Question 4

Finally, select four countries from one continent and create visualizations for the daily number of confirmed cases per 100,000 and the daily number of deaths per 100,000 people between March 15, 2020, and December 31, 2021. 

```{r p3q4-response}
country = "Philippines"

country_cases <- csse_global_cases_longer %>%
  filter(Country_Name == country & date >= "2020-03-15" & date < "2022-01-01")

country_deaths <- csse_global_deaths_longer %>%
  filter(Country_Name == country & date >= "2020-03-15" & date < "2022-01-01") %>%
  select(-"Country_Name")

country_cases
country_deaths

country_combined <- country_cases %>%
  left_join(country_deaths, by = c("date" = "date"))

country_2020_po <- global_population_estimates %>%
  filter(`Country Name` == country) %>%
  select(`2020 [YR2020]`)

country_2021_po <- global_population_estimates %>%
  filter(`Country Name` == country) %>%
  select(`2021 [YR2021]`)

country_2020_pop <- as.numeric(country_2020_po)
country_2021_pop <- as.numeric(country_2021_po)


country_2020_pop
country_2021_pop
country_cleaned <- country_combined %>%
  mutate(population = case_when(
    grepl("2020", date) ~ country_2020_pop,
    grepl("2021", date) ~ country_2021_pop
  ))

country_cleaned_100k <- country_cleaned %>%
  mutate(
    cases_100k = cases/population*100000,
    deaths_100k = deaths/population*100000
  )

philippines_country <- country_cleaned_100k %>%
  select(Country_Name, date, cases_100k, deaths_100k)

philippines_country

#----------------------------------------------------------------------------------------------------------------

country = "Singapore"

country_cases <- csse_global_cases_longer %>%
  filter(Country_Name == country & date >= "2020-03-15" & date < "2022-01-01")

country_deaths <- csse_global_deaths_longer %>%
  filter(Country_Name == country & date >= "2020-03-15" & date < "2022-01-01") %>%
  select(-"Country_Name")

country_cases
country_deaths

country_combined <- country_cases %>%
  left_join(country_deaths, by = c("date" = "date"))

country_combined

country_2020_po <- global_population_estimates %>%
  filter(`Country Name` == country) %>%
  select(`2020 [YR2020]`)

country_2021_po <- global_population_estimates %>%
  filter(`Country Name` == country) %>%
  select(`2021 [YR2021]`)

country_2020_pop <- as.numeric(country_2020_po)
country_2021_pop <- as.numeric(country_2021_po)


country_2020_pop
country_2021_pop
country_cleaned <- country_combined %>%
  mutate(population = case_when(
    grepl("2020", date) ~ country_2020_pop,
    grepl("2021", date) ~ country_2021_pop
  ))

country_cleaned

country_cleaned_100k <- country_cleaned %>%
  mutate(
    cases_100k = cases/population*100000,
    deaths_100k = deaths/population*100000
  )

country_cleaned_100k

singapore_country <- country_cleaned_100k %>%
  select(Country_Name, date, cases_100k, deaths_100k)

singapore_country

#----------------------------------------------------------------------------------------------------------------

country = "Japan"

country_cases <- csse_global_cases_longer %>%
  filter(Country_Name == country & date >= "2020-03-15" & date < "2022-01-01")

country_deaths <- csse_global_deaths_longer %>%
  filter(Country_Name == country & date >= "2020-03-15" & date < "2022-01-01") %>%
  select(-"Country_Name")

country_cases
country_deaths

country_combined <- country_cases %>%
  left_join(country_deaths, by = c("date" = "date"))

country_combined

country_2020_po <- global_population_estimates %>%
  filter(`Country Name` == country) %>%
  select(`2020 [YR2020]`)

country_2021_po <- global_population_estimates %>%
  filter(`Country Name` == country) %>%
  select(`2021 [YR2021]`)

country_2020_pop <- as.numeric(country_2020_po)
country_2021_pop <- as.numeric(country_2021_po)


country_2020_pop
country_2021_pop
country_cleaned <- country_combined %>%
  mutate(population = case_when(
    grepl("2020", date) ~ country_2020_pop,
    grepl("2021", date) ~ country_2021_pop
  ))

country_cleaned

country_cleaned_100k <- country_cleaned %>%
  mutate(
    cases_100k = cases/population*100000,
    deaths_100k = deaths/population*100000
  )

country_cleaned_100k

japan_country <- country_cleaned_100k %>%
  select(Country_Name, date, cases_100k, deaths_100k)

japan_country

#----------------------------------------------------------------------------------------------------------------

country = "Vietnam"

country_cases <- csse_global_cases_longer %>%
  filter(Country_Name == country & date >= "2020-03-15" & date < "2022-01-01")

country_deaths <- csse_global_deaths_longer %>%
  filter(Country_Name == country & date >= "2020-03-15" & date < "2022-01-01") %>%
  select(-"Country_Name")

country_cases
country_deaths

country_combined <- country_cases %>%
  left_join(country_deaths, by = c("date" = "date"))

country_combined

country_2020_po <- global_population_estimates %>%
  filter(`Country Name` == country) %>%
  select(`2020 [YR2020]`)

country_2021_po <- global_population_estimates %>%
  filter(`Country Name` == country) %>%
  select(`2021 [YR2021]`)

country_2020_pop <- as.numeric(country_2020_po)
country_2021_pop <- as.numeric(country_2021_po)


country_2020_pop
country_2021_pop
country_cleaned <- country_combined %>%
  mutate(population = case_when(
    grepl("2020", date) ~ country_2020_pop,
    grepl("2021", date) ~ country_2021_pop
  ))

country_cleaned

country_cleaned_100k <- country_cleaned %>%
  mutate(
    cases_100k = round(cases/population*100000, 2),
    deaths_100k = round(deaths/population*100000, 2)
  )

country_cleaned_100k

vietnam_country <- country_cleaned_100k %>%
  select(Country_Name, date, cases_100k, deaths_100k)

vietnam_country


#--------------------------------------------------------------------------------

ultimate_combined <- bind_rows(philippines_country, singapore_country, japan_country, vietnam_country)
ultimate_combined

ggplot(data = ultimate_combined, aes(x=date, y=cases_100k)) +
  geom_line(aes(color = Country_Name))

ggplot(data = ultimate_combined, aes(x=date, y=deaths_100k)) +
  geom_line(aes(color = Country_Name))

print("In terms of cases per 100k in population, Singapore has the biggest rate, which means on average per person, it has the most cases.
      Though this is the case, Philippines has the highest deaths per 100k population, which means its cases count is acceptable, but
      it is not managing its death rate well.")
```

-- Communicate your methodology, results, and interpretation here -- 